---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BayesOmics

<!-- badges: start -->
<!-- badges: end -->

BayesOmics is a Bayesian differential analysis method for omics data that involve capturing the correlation structure through a kernel function. In this approach, users are required to select an appropriate kernel, optimize its hyperparameters, and subsequently utilize the correlation matrix derived from this kernel to compute the posterior.

## See it in action

Given a long-format dataset (one row per id/group/sample), BayesOmics fits a single
kernel-structured posterior per group and turns it into a constant-size overlap
heatmap and a per-id posterior-mean profile, however many groups are present:

```{r message=FALSE, warning=FALSE}
library(keRnel)
library(BayesOmics)

set.seed(42)
data <- simu_db(nb_id = 25, nb_group = 4, nb_sample = 3, diff_group = 8)

kern <- new("SEKernel")
kern <- set_hyperparameters(kern, c(1.0, 1.0))
opt  <- optim_hp(c(1.0, 1.0), data, rep(0, nrow(data)), kern, diag(0.1, nrow(data)))
kern <- set_hyperparameters(kern, opt)

posterior <- multi_posterior_mean(data, kern)
samples   <- sample_posterior(posterior, n = 2000)
```

```{r}
plot_group_overlap_heatmap(samples, id = unique(samples$ID)[1])
```

```{r}
plot_posterior_mean(samples)
```

## Installation

You can install the development version of BayesOmics from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("terenceviellard/BayesOmics")
```


## Bayes Omics in a nutshell

This is a basic example which shows you how to solve a common problem:

Generate a synthetic dataset with the correct format:

```{r}
set.seed(123)
data <- simu_db(nb_id = 400,
                nb_group = 2,
                nb_sample = 1)
head(data)
```


We choose a Squared Exponential (SE) kernel, given by:

$$
K_{\text{SE}}(x, x') = \sigma^2 \exp\left(-\frac{\|x - x'\|^2}{2\ell^2}\right)
$$

where:

- \( K_{\text{SE}}(x, x') \) is the covariance between inputs \( x \) and \( x' \)

- \( \sigma^2 \) is the signal variance.

- \( \ell \) is the length scale.

- \( \|x - x'\| \) is the Euclidean distance between \( x \) and \( x' \).

```{r}
SEKernel <- new("SEKernel")
```

We initialize the hyperparameters with \(\sigma^2\) = 1 and \(\ell\) = 1.

```{r}
hp <- c(1.0, 1.0)
SEKernel <- set_hyperparameters(SEKernel, hp)
```

We set identical priors for the different groups.

```{r}
n <- dim(data)[1]
prior_mean_nul <- rep(0, n)
prior_cov_nul <- diag(0.1, n)
```


Then we need to optimize  \( \sigma^2 \)  and  \( \ell \) :

```{r}
opt= optim_hp(
  hp = hp,
  db = data,
  prior_mean = prior_mean_nul,
  kern = SEKernel,
  prior_cov = prior_cov_nul
)
opt
```

We create a new kernel with optimal hyperparameters.
```{r}
SEKernelopt <- set_hyperparameters(SEKernel, opt)
```


The parameters of all posterior distributions can be computed. Each group's posterior
mean follows a Normal-Normal conjugate update:

$$
p(\mathbf{\mu} \mid y_1, \dots, y_N, \Sigma_{\hat{\theta}}) = \mathcal{N}\left(\mathbf{\mu}; \ \dfrac{\lambda_0 \mu_0 + \sum_{n=1}^{N} y_n}{N + \lambda_0}, \dfrac{1}{N + \lambda_0} \Sigma_{\hat{\theta}}\right)
$$

where \( \Sigma_{\hat{\theta}} \) is the correlation matrix built from the optimized kernel,
\( \mu_0 \) and \( \lambda_0 \) are the prior mean and prior precision, and \( N \) is the
number of observations.

```{r}
posterior <- multi_posterior_mean(data, SEKernelopt)
```

Once parameters of the posterior distributions are available, we can calculate the
overlapping coefficient (OVL) between the two groups, used here as our differential
analysis statistic:

$$
OVL = 2 \Phi\!\left(-\frac{\delta}{2}\right)
$$

where \( \delta \) is the Mahalanobis distance between the two groups' posterior means
under their pooled covariance, and \( \Phi \) is the standard normal CDF. \( OVL = 1 \)
means the two posterior distributions are identical (no differential signal); \( OVL
\approx 0 \) means they are fully separated (strong differential signal).

```{r}
calculate_group_overlaps(posterior)
```

## Visualizing results

Once a posterior is computed, draw samples from it with `sample_posterior()` and feed
them to any of the `plot_*()` functions below. They all share the same long-format
input (`ID`, `Group`, `Sample`), so they compose freely.

```{r}
samples <- sample_posterior(posterior, n = 2000)
```

`plot_posterior_overlap()` overlays the two groups' posterior densities for one id and
shades the probability mass used to compute the OVL coefficient above:

```{r}
plot_posterior_overlap(samples, id = unique(samples$ID)[1])
```

`plot_posterior_mean()` gives a region-wide view: one point per id, coloured by group,
so you can spot which ids drive the differential signal:

```{r}
plot_posterior_mean(samples)
```

`plot_group_overlap_heatmap()` summarizes every pairwise group comparison for one id as
a single tile grid -- the same plot already shown above, useful here as it stays
readable regardless of how many groups are compared:

```{r}
plot_group_overlap_heatmap(samples, id = unique(samples$ID)[1])
```

## Bayes Omics for multigroup analysis

We can use Bayes Omics to compare more than 2 groups.
Generate a synthetic dataset with 3 groups with the correct format:


```{r}
data <- simu_db(nb_id = 200,
                nb_group = 3,
                nb_sample = 1)
```


```{r}
SEKernel <- new("SEKernel")

hp <- c(1.0, 1.0)
SEKernel <- set_hyperparameters(SEKernel, hp)

n <- dim(data)[1]

prior_mean_nul <- rep(0, n)
prior_cov_nul <- diag(0.1, n)

opt <- optim_hp(
  hp = hp,
  db = data,
  prior_mean = prior_mean_nul,
  kern = SEKernel,
  prior_cov = prior_cov_nul
)
opt
```

We create a new kernel with optimal hyperparameters.
```{r}
SEKernelopt <- set_hyperparameters(SEKernel, opt)
```


The parameters of the posterior distributions for the 3 groups can be computed using
the same Normal-Normal conjugate update as above:

```{r}
posterior <- multi_posterior_mean(data, SEKernelopt)
```

Once parameters of the posterior distributions are available, `calculate_group_overlaps()`
returns a full pairwise OVL matrix, with one row/column per group:

```{r}
calculate_group_overlaps(posterior)
```

As with the 2-group case, `plot_group_overlap_heatmap()` gives a constant-size summary
of every pair at once -- unlike a full grid of one density panel per pair, which grows
as the number of groups squared and quickly becomes illegible:

```{r}
samples <- sample_posterior(posterior, n = 2000)
plot_group_overlap_heatmap(samples, id = unique(samples$ID)[1])
```

## Go further

| Vignette | What you will find |
|---|---|
| `vignette("BayesOmics")` | Step-by-step model walkthrough: kernel selection, hyperparameter optimization, posterior computation, and every plot function, on both a two-group and a multi-group dose-response example. |
| `vignette("sample_size_scenarios")` | How unbalanced designs (different replicate counts per group) behave: each group's posterior precision depends only on its own sample size, and the comparison remains valid even with strongly unequal counts. |
| `vignette("troubleshooting")` | Exact error messages you may encounter, what causes each of them, and the targeted fix. |
