---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BayesOmics

<!-- badges: start -->
<!-- badges: end -->

Bayesian differential analysis for omics data.The correlation structure is captured by a kernel. Users need to choose a kernel, optimize the hyperparameters, and then use the correlation matrix computed by this kernel in the analysis.
## Installation

You can install the development version of BayesOmics from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("terenceviellard/BayesOmics")
```

## Example

This is a basic example which shows you how to solve a common problem:

### DATA 
```{r}
library(BayesOmics)
```

```{r}
data <- simu_db()
head(data)
```
```{r}
colnames(data)
```
You need to have 6 columns in long format like this


### Kernel 

First, choose a kernel, in our example, kernel SE defined as:

$$
k_{\text{SE}}(x, x') = \sigma^2 \exp\left(-\frac{\ x - x'\|^2}{2\ell^2}\right)
$$

where:
- \( \sigma^2 \) is the signal variance,
- \( \ell \) is the length scale,
- \( x \) and \( x' \) are input vectors.

Then we need to optimize Hps  \( \sigma^2 \)  and  \( \ell \) : 

```{r eval=FALSE, include=TRUE}
SEKernel <- new("SEKernel")

hp <- c(1.0, 1.0)
SEKernel <- set_hyperparameters(SEKernel, hp)

n <- dim(data)[1]

mean_nul <- rep(0, n)
post_cov_nul <- diag(0.1, n)
pen_diag <- 1e-6



opt <- optim_hp(
  hp = hp,
  db = df,
  mean = mean_nul,
  kern = SEKernel,
  post_cov = post_cov_nul,
  pen_diag = pen_diag
)
opt$par
```

We create a new kernel with optimal HpS
```{r}
# En attendant de reparer optimHP :
SEKernel <- new("SEKernel")
opt <- list()
opt$par <- c(1, 1)
SEKernelopt <- set_hyperparameters(SEKernel, opt$par)
```


### Posterior 

```{r}
posterior <- multi_posterior_mean(data, SEKernelopt)
```

### DIfferential analtsis 

```{r}
calculate_group_overlaps(posterior)
```

