---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BayesOmics

<!-- badges: start -->
<!-- badges: end -->

BayesOmics is an R package for Bayesian differential analysis of omics data.
It places a kernel over the feature dimension (genomic position, mass-to-charge
ratio, retention time, …) to capture correlation structure, fits that kernel by
maximum likelihood, and computes a posterior profile per group. Groups are then
compared as complete multivariate profiles via an overlapping coefficient (OVL),
giving a single, interpretable differential-analysis statistic per region.

## When to use BayesOmics

BayesOmics is designed for datasets where features are not independent
but share a meaningful ordering or distance along some axis — genomic position,
retention time, dose level, time point, m/z ratio, or any other continuous
covariate. Classical per-feature tests (t-test, limma) ignore that structure
and multiply the number of comparisons. BayesOmics instead models the whole
region jointly: it fits a kernel over the feature axis to capture the
correlation pattern, then derives one posterior profile per group and compares
groups as complete multivariate objects via a single overlapping coefficient
(OVL).

Typical use cases:

- **Longitudinal / time-course** omics: features ordered by time point or dose
- **Spatial genomics**: features ordered by chromosomal position (methylation,
  ChIP-seq, ATAC-seq)
- Any experiment where the signal is expected to vary smoothly across the
  feature axis

## Key assumptions

- Every group must be measured at the **same set of Input values** (same feature
  positions) so that groups share one kernel matrix and can be compared.
- Within a group, every feature (`ID`) must have the **same number of
  replicates** (`Sample`).
- Groups being compared must cover the **same set of IDs**.Groups may have different numbers of replicates.

See `vignette("troubleshooting")` for the exact error raised when these
conditions are not met.

## Installation

```r
# install.packages("pak")
pak::pak("terenceviellard/BayesOmics")
```

## Dependencies

BayesOmics depends on the companion package
[**keRnel**](https://github.com/terenceviellard/keRnel) for kernel classes,
pairwise kernel computation, and hyperparameter management.
Both packages are installed together via `pak::pak("terenceviellard/BayesOmics")`.

CRAN dependencies: `mvtnorm`, `ggplot2`, `dplyr`, `gridExtra`.

## See it in action

Five lines from raw data to posterior comparison:

```{r message=FALSE, warning=FALSE}
library(keRnel)
library(BayesOmics)

set.seed(42)
data <- simu_db(nb_id = 25, nb_group = 4, nb_sample = 3, diff_group = 8)

kern         <- variance_kernel(variance = 1) * se_kernel(length_scale = 1)
control_data <- data[data$Group == 1, ]
opt          <- optim_hp(kern, control_data,
                         prior_mean = mean(control_data$Output), prior_cov = 1)
kern         <- do.call(kupdate, c(list(kern), as.list(opt)))

posterior <- multi_posterior_mean(data, kern)
calculate_group_overlaps(posterior)
samples   <- sample_posterior(posterior, n = 2000)
```

Pairwise OVL between every group for one feature (lower = more differentiated):

```{r}
plot_group_overlap_heatmap(samples, id = unique(samples$ID)[1])
```

Posterior mean at every feature, coloured by group:

```{r}
plot_posterior_mean(samples)
```

## Go further

| Vignette | What you will find |
|---|---|
| `vignette("get-started")` | Data format, kernel choice, step-by-step pipeline, and basic plots — the right place to start. |
| `vignette("omics-analysis")` | Full worked examples (two-group and multi-group dose-response) with kernel composition, HP interpretation, and every plot function. |
| `vignette("sample_size_scenarios")` | How unbalanced designs affect posterior width and the OVL, and why the comparison remains valid with unequal sample sizes. |
| `vignette("troubleshooting")` | Exact error messages you may encounter, what causes each of them, and the targeted fix. |
