---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BayesOmics

<!-- badges: start -->
<!-- badges: end -->


BayesOmics is a Bayesian differential analysis method for omics data that involve capturing the correlation structure through a kernel function. In this approach, users are required to select an appropriate kernel, optimize its hyperparameters, and subsequently utilize the correlation matrix derived from this kernel to compute the posterior.

## Installation

You can install the development version of BayesOmics from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("terenceviellard/BayesOmics")
```

## Bayes Omics in a nutshell

This is a basic example which shows you how to solve a common problem:

```{r}
library(BayesOmics)
```
Generate a synthetic dataset with the correct format:

```{r}
set.seed(123)
data <- simu_db(nb_id = 400,
                nb_group = 2,
                nb_sample = 1)
head(data)
```


We choose a Squared Exponential (SE) kernel, given by:

$$
K_{\text{SE}}(x, x') = \sigma^2 \exp\left(-\frac{\|x - x'\|^2}{2\ell^2}\right)
$$

where:

- \( K_{\text{SE}}(x, x') \) is the covariance between inputs \( x \) and \( x' \)

- \( \sigma^2 \) is the signal variance.

- \( \ell \) is the length scale.

- \( \|x - x'\| \) is the Euclidean distance between \( x \) and \( x' \).

```{r}
SEKernel <- new("SEKernel")
```

We initialize the hyperparameters with \(\sigma^2\) = 1 and \(\ell\) = 1.

```{r}
hp <- c(1.0, 1.0)
SEKernel <- set_hyperparameters(SEKernel, hp)
```

We set identical priors for the different groups.

```{r}
n <- dim(data)[1]
mean_nul <- rep(0, n)
post_cov_nul <- diag(0.1, n)
```


Then we need to optimize  \( \sigma^2 \)  and  \( \ell \) : 

```{r}
opt= optim_hp(
  hp = hp,
  db = data,
  mean = mean_nul,
  kern = SEKernel,
  post_cov = post_cov_nul
)
opt
```

We create a new kernel with optimal hyperparameters.
```{r}
SEKernelopt <- set_hyperparameters(SEKernel, opt)
```


The parameters of all posterior distributions can be computed thanks to:

```{r}
posterior <- multi_posterior_mean(data, SEKernelopt)
```

Once parameters of the posterior distributions are available, we can calculate the overlapping coefficient for our differential analysis.
```{r}
calculate_group_overlaps(posterior)
```


## Bayes Omics for multigroup analysis 

We can use Bayes Omics to compare more than 2 groups.
Generate a synthetic dataset with 5 groups with the correct format:


```{r}
data <- simu_db(nb_id = 200,
                nb_group = 5,
                nb_sample = 1)
```


```{r}
SEKernel <- new("SEKernel")

hp <- c(1.0, 1.0)
SEKernel <- set_hyperparameters(SEKernel, hp)

n <- dim(data)[1]

mean_nul <- rep(0, n)
post_cov_nul <- diag(0.1, n)

opt <- optim_hp(
  hp = hp,
  db = data,
  mean = mean_nul,
  kern = SEKernel,
  post_cov = post_cov_nul
)
opt
```

We create a new kernel with optimal hyperparameters.
```{r}
SEKernelopt <- set_hyperparameters(SEKernel, opt)
```


The parameters of the posterior distributions for the 5 groups can be computed using:
```{r}
posterior <- multi_posterior_mean(data, SEKernelopt)
```

Once parameters of the posterior distributions are available. 
```{r}
calculate_group_overlaps(posterior)
```
