---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# BayesOmics

<!-- badges: start -->
<!-- badges: end -->

Bayesian differential analysis for omics data.The correlation structure is captured by a kernel. Users need to choose a kernel, optimize the hyperparameters, and then use the correlation matrix computed by this kernel in the analysis.
## Installation

You can install the development version of BayesOmics from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("terenceviellard/BayesOmics")
```

## Bayes Omics in a nutshell

This is a basic example which shows you how to solve a common problem:

```{r}
library(BayesOmics)
```
Generate a synthetic dataset with the correct format:

```{r}
set.seed(123)
data <- simu_db(nb_id = 400,
                nb_group = 2,
                nb_sample = 1)
head(data)
```

```{r}
colnames(data)
```
You need to have 5 columns in long format like this

We choose a Squared Exponential (SE) kernel, given by:

$$
K_{\text{SE}}(x, x') = \sigma^2 \exp\left(-\frac{\|x - x'\|^2}{2\ell^2}\right)
$$

where:
- \( K_{\text{SE}}(x, x') \) is the covariance between inputs \( x \) and \( x' \).
- \( \sigma^2 \) is the signal variance.
- \( \ell \) is the length scale.
- \( \|x - x'\| \) is the Euclidean distance between \( x \) and \( x' \).
Then we need to optimize  \( \sigma^2 \)  and  \( \ell \) : 

```{r}
SEKernel <- new("SEKernel")

hp <- c(1.0, 1.0)
SEKernel <- set_hyperparameters(SEKernel, hp)

n <- dim(data)[1]

mean_nul <- rep(0, n)
post_cov_nul <- diag(0.1, n)

opt <- optim_hp(
  hp = hp,
  db = data,
  mean = mean_nul,
  kern = SEKernel,
  post_cov = post_cov_nul
)
opt
```

We create a new kernel with optimal HpS
```{r}
SEKernelopt <- set_hyperparameters(SEKernel, opt)
```


The parameters of all posterior distributions can be computed thanks to:

```{r}
posterior <- multi_posterior_mean(data, SEKernelopt)
```

Once parameters of the posterior distributions are available, we can calculate the overlapping coefficient for our differential analysis.
```{r}
calculate_group_overlaps(posterior)
```


## Bayes Omics for multigroup analysis 

We can use Bayes Omics to compare more than 2 groups.
Generate a synthetic dataset with 5 groups with the correct format:


```{r}
data <- simu_db(nb_id = 200,
                nb_group = 5,
                nb_sample = 1)
```

We also choose a SEKernel so we need to optimize Hps  \( \sigma^2 \)  and  \( \ell \) : 

```{r}
SEKernel <- new("SEKernel")

hp <- c(1.0, 1.0)
SEKernel <- set_hyperparameters(SEKernel, hp)

n <- dim(data)[1]

mean_nul <- rep(0, n)
post_cov_nul <- diag(0.1, n)

opt <- optim_hp(
  hp = hp,
  db = data,
  mean = mean_nul,
  kern = SEKernel,
  post_cov = post_cov_nul
)
opt
```

We create a new kernel with optimal HpS
```{r}
SEKernelopt <- set_hyperparameters(SEKernel, opt)
```


The parameters of all posterior distributions can be computed thanks to:

```{r}
posterior <- multi_posterior_mean(data, SEKernelopt)
```

Once parameters of the posterior distributions are available. 
```{r}
library(bayestestR)
calculate_group_overlaps(posterior)
```
