---
title: "Basic pipeline: comparing two groups with a spatial kernel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic pipeline: comparing two groups with a spatial kernel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup, message = FALSE, warning = FALSE}
library(keRnel)
library(BayesOmics)
library(ggplot2)
```

You have two groups, one covariate (`Input`), for instance time,  per feature, and want to know
whether -- and where -- the groups differ. This is the shortest path through
the full pipeline: simulate, fit, compare, visualize.

## Simulate

```{r}
set.seed(1)
kern_true <- variance_kernel(variance = 10) * se_kernel(length_scale = 100)
data <- simu_db_kernel(
  kernel        = kern_true,
  nb_id         = 6,
  nb_group      = 2,
  nb_sample     = 8,
  diff_group    = 3,
  var_sample    = 3,
  range_input   = c(0, 200),
  integer_input = TRUE,
  input_grid    = TRUE
)
head(data)
```

```{r}
ggplot(data, aes(Input, Output, color = Group, group = interaction(Group, Sample))) +
  geom_line(alpha = 0.4) +
  geom_point(size = 1)
```

Each line is one replicate; the two groups' vertical offset is the
differential signal the rest of this example tries to recover and quantify.

## Fit the kernel

A Squared Exponential (SE) term captures spatial correlation; a white-noise
term absorbs measurement noise and keeps the resulting covariance
well-conditioned (see the kernel-choice article for other kernels and why the
noise term matters):

```{r}
kern <- variance_kernel(variance = 1) * se_kernel(length_scale = 50) +
  white_noise_kernel(noise = 1)

group1     <- data[data$Group == unique(data$Group)[1], ]
prior_mean <- mean(group1$Output)
hp_opt     <- fit_kernel(kern, group1, prior_mean = prior_mean, prior_cov = 1)
kern_opt   <- kupdate(kern,
                       variance     = hp_opt[["variance"]],
                       length_scale = hp_opt[["length_scale"]],
                       noise        = hp_opt[["noise"]])
```

## Compute posteriors and compare

```{r}
posterior <- posterior_mean(data, kern_opt, mu_0 = prior_mean, lambda_0 = 1)
group_diff(posterior)
```

The Overlapping Coefficient (OVL) ranges from 1 (identical posteriors, no
differential signal) to 0 (fully separated, strong differential signal).

## Visualize one feature

```{r}
samples <- sample_posterior(posterior, n = 2000)
plot_posterior_overlap(samples,
                        group1 = unique(samples$Group)[1],
                        group2 = unique(samples$Group)[2],
                        id     = unique(samples$ID)[1])
```

The shaded region is exactly what the OVL number above quantifies for that
feature.

## How many features differ?

The single-feature view above does not say whether the difference is
localized or spread across the whole region. `compute_multi_diff()` answers
that from the posterior draws: the empirical probability that group 1
exceeds group 2 on `k` out of the 6 features, plotted by `plot_multi_diff()`
alongside every feature's posterior mean:

```{r}
multi_diff <- compute_multi_diff(samples, results = posterior)
plot_multi_diff(multi_diff)
```

The bar chart is concentrated almost entirely at `k = 0` (over 98% of the
probability mass), not spread uniformly across `0`-`6`: on almost every
feature, group 2's draw exceeds group 1's. This is what a clear, consistent
differential signal looks like -- not just a low joint OVL, but the same
direction of effect on nearly every feature.

## Takeaways

- The pipeline is always four calls: `fit_kernel()` -> `posterior_mean()` ->
  `group_diff()` -> a plot function.
- This example fixes every other variable: two groups, one covariate, a
  kernel shared across groups. Each is relaxed in a separate article.

## Related

*Univariate mode* (no covariate at all) -- *Pooled vs. non-pooled fitting*
(one kernel per group) -- *Comparing more than two groups*.
