---
title: "Univariate mode: comparing two groups with no feature axis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Univariate mode: comparing two groups with no feature axis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

```{r setup, message = FALSE}
library(BayesOmics)
library(ggplot2)
```

You have two groups of replicate measurements on a single quantity -- no
feature/covariate axis at all (e.g. one protein's abundance, or a single
summary score). There is nothing for a kernel to correlate against, so
supplying neither `ID` nor `Input` switches `posterior_mean()` into
univariate mode: each group is treated as one feature.

## Data

```{r}
set.seed(1)
data <- simu_db(univariate = TRUE, nb_group = 2, nb_sample = 12,
                 diff_group = 1.5, var_sample = 4)
data$Group <- factor(data$Group, levels = c(1, 2), labels = c("Control", "Treatment"))
head(data)
```

```{r}
ggplot(data, aes(Group, Output, color = Group)) +
  geom_jitter(width = 0.1, size = 2)
```

## Fit and compare

No `kern` is supplied either: with a single feature per group,
`posterior_mean()` fits a closed-form residual-variance kernel itself.

```{r}
posterior <- posterior_mean(data, mu_0 = mean(data$Output), lambda_0 = 1)
group_diff(posterior)
```

The warning above ("running in univariate mode...") is expected -- it
confirms the mode switch rather than signalling a mistake.

## Visualize the difference

With a single feature per group, `plot_posterior_overlap()` shows exactly
what the OVL number summarizes: both groups' posterior densities, with the
region of overlap shaded:

```{r}
samples <- sample_posterior(posterior, n = 2000)
plot_posterior_overlap(samples, group1 = "Control", group2 = "Treatment",
                        id = unique(samples$ID)[1])
```

## Takeaways

- Univariate mode needs both `ID` and `Input` *absent*, not just one of them.
- Everything downstream (`group_diff()`, the metric family,
  the plots) works unchanged -- a single feature is just a group with `nb_id = 1`.

## Related

*Basic pipeline* (the general multi-feature case) -- *Pooled vs. non-pooled
fitting*.
