---
title: "Pooled vs. non-pooled: one kernel for every group, or one per group"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pooled vs. non-pooled: one kernel for every group, or one per group}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

```{r setup, message = FALSE}
library(BayesOmics)
library(keRnel)
library(ggplot2)
```

`fit_kernel(..., group_col = "Group", pooled = )` decides whether a kernel
is fit once, shared across every group (`pooled = TRUE`, the default), or
independently, one fit per group (`pooled = FALSE`). Here, `Treatment` has a
genuinely *larger kernel variance* than `Control` -- still the same shared
`length_scale`.
The rest of the framework is agnostic to this choice, whether in univariate or multivariate analyses. For analyses with few replicates, the “pooled” option is preferred to improve the quality of the kernel estimate. 
## Data

```{r}
set.seed(2)
nb_id <- 15
ids <- paste0("ID_", seq_len(nb_id))
positions <- matrix(sort(runif(nb_id, 0, 30)), ncol = 1)
rownames(positions) <- ids

kern_control   <- variance_kernel(variance = 1) * se_kernel(length_scale = 5)
kern_treatment <- variance_kernel(variance = 4) * se_kernel(length_scale = 5)

make_group_kernel <- function(kern_true, label, nb_sample = 10, var_sample = 0.3) {
  Sigma <- evaluate(kern_true, positions, positions) + var_sample * diag(nb_id)
  L <- t(chol(Sigma))
  do.call(rbind, lapply(seq_len(nb_sample), function(s) {
    y <- 10 + as.vector(L %*% rnorm(nb_id))
    data.frame(ID = ids, Group = label, Sample = s, Input = positions[, 1], Output = y)
  }))
}
data <- rbind(
  make_group_kernel(kern_control,   "Control"),
  make_group_kernel(kern_treatment, "Treatment")
)
```

## Fit both ways

```{r}
kern_template <- variance_kernel(variance = 1) * se_kernel(length_scale = 1)

kern_pooled <- fit_kernel(kern_template, data, prior_cov = 0.3,
                           group_col = "Group", pooled = TRUE, verbose = TRUE)$kern

# pooled = FALSE fits each group directly on its own (non-demeaned) data, so
# -- unlike the pooled fit above, where group_col's demeaning makes prior_mean
# irrelevant -- prior_mean must match Output's real baseline here (10).
fits_nonpooled <- fit_kernel(kern_template, data, prior_mean = 10, prior_cov = 0.3,
                              group_col = "Group", pooled = FALSE, verbose = TRUE)
kern_by_group  <- lapply(fits_nonpooled, function(f) f$kern)

get_trainable_params(kern_pooled)
lapply(kern_by_group, get_trainable_params)
```

`pooled = FALSE` recovers each group's own `variance` (close to the true `1`
and `4`); `pooled = TRUE` compromises on a single value in between, since it
assumes one kernel is shared.

## Visualizing the fitted kernels

Since these are `variance * se_kernel(length_scale)` kernels, their shape is
fully described by the covariance between two Input positions as a function
of the distance between them. Plotting that curve for the pooled kernel and
for each group's own non-pooled kernel shows exactly what "compromise" means
here: the pooled curve sits between the two, in both height (variance) and
spread (length_scale):

```{r}
dist_grid <- seq(0, 15, length.out = 100)
ref      <- matrix(0, nrow = 1)
grid_mat <- matrix(dist_grid, ncol = 1)

curve_df <- rbind(
  data.frame(distance = dist_grid,
             covariance = as.vector(evaluate(kern_pooled, ref, grid_mat)),
             Kernel = "Pooled (shared)"),
  data.frame(distance = dist_grid,
             covariance = as.vector(evaluate(kern_by_group$Control, ref, grid_mat)),
             Kernel = "Control (non-pooled)"),
  data.frame(distance = dist_grid,
             covariance = as.vector(evaluate(kern_by_group$Treatment, ref, grid_mat)),
             Kernel = "Treatment (non-pooled)")
)

ggplot(curve_df, aes(distance, covariance, color = Kernel)) +
  geom_line(linewidth = 1) +
  labs(x = "Distance between Input positions", y = "Kernel covariance",
       title = "Fitted kernel shape: pooled vs. non-pooled")
```


## Takeaways

- `pooled` in `fit_kernel(..., group_col = "Group", pooled = )` means the
  same thing as in `posterior_mean(kern = NULL, pooled = )` for the
  closed-form fit: one fit shared across every group, or one independent
  fit per group.
- Non-pooled fitting is more faithful when groups genuinely differ (in
  noise, or in the kernel's own hyperparameters). `group_diff()`/`ovl_metric()`
  still work in that case -- they fall back to a Monte Carlo/KDE estimate
  automatically instead of the exact closed form (see the metric-choice
  article for the accuracy tradeoff of that fallback).

## Related

*Choice of metric: OVL vs. Wasserstein* -- *Basic pipeline*.
