---
title: "Comparing more than two groups"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing more than two groups}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

```{r setup, message = FALSE, warning = FALSE}
library(keRnel)
library(BayesOmics)
library(ggplot2)
```

Nothing in the pipeline changes when there are more than two groups: every
function below is the same call as in the two-group case. What changes is
how to read the result -- a `g x g` matrix instead of a single number, and a
compact plot to go with it. A four-arm dose-response design, where two doses
happen to produce the same effect:

## Simulate and fit

```{r}
set.seed(2)
kern_true <- variance_kernel(variance = 10) * se_kernel(length_scale = 80)
data <- simu_db_kernel(
  kernel        = kern_true,
  nb_id         = 6,
  nb_group      = 4,
  nb_sample     = 8,
  diff_group    = c(0, 5, 5, 10),
  var_sample    = 3,
  range_input   = c(0, 200),
  group_labels  = c("Control", "Dose1", "Dose2", "Dose3"),
  integer_input = TRUE,
  input_grid    = TRUE
)

kern <- variance_kernel(variance = 1) * se_kernel(length_scale = 50) +
  white_noise_kernel(noise = 1)
g1         <- data[data$Group == "Control", ]
prior_mean <- mean(g1$Output)
hp_opt     <- fit_kernel(kern, g1, prior_mean = prior_mean, prior_cov = 1)
kern_opt   <- kupdate(kern,
                       variance     = hp_opt[["variance"]],
                       length_scale = hp_opt[["length_scale"]],
                       noise        = hp_opt[["noise"]])
posterior  <- posterior_mean(data, kern_opt, mu_0 = prior_mean, lambda_0 = 1)
```

```{r}
ggplot(data, aes(Input, Output, color = Group, group = interaction(Group, Sample))) +
  geom_line(alpha = 0.4) +
  geom_point(size = 1)
```

`Dose1` and `Dose2` visually overlap across most of the region, while
`Control` and `Dose3` sit clearly apart -- the same pattern the pairwise
matrix below quantifies.

## Read the pairwise matrix

```{r}
group_diff(posterior)
```

The matrix is symmetric with 1s on the diagonal. Read it row/column by
row/column: `Dose1`-`Dose2` stands out as the least separated off-diagonal
pair (OVL ~ 0.064, clearly the largest off-diagonal value), consistent
with both having the same simulated effect (`diff_group = 5`) -- every
other pair is well separated.

## One-feature heatmap

For a single feature, the same information is a small `g x g` tile plot --
useful once you want to compare many features' matrices side by side rather
than reading numbers:

```{r}
samples <- sample_posterior(posterior, n = 2000)
plot_group_overlap_heatmap(samples, id = unique(samples$ID)[1])
```

## Every pairwise comparison at once

`plot_multi_diff()` is not capped at two groups either. With more than two
groups it shows an at-a-glance overview immediately (a heatmap, one tile
per pair, plus the per-feature posterior-mean panel) and returns every
individual pair's full panel for inspection on demand -- no cap, never
crammed into one shared grid:

```{r}
multi_diff <- compute_multi_diff(samples, results = posterior)
out <- plot_multi_diff(multi_diff)  # displays the heatmap + mean panel above
```

```{r}
out$pairs[["Dose1_vs_Dose2"]]
```

`Dose1`-`Dose2` is the one panel closest to a uniform spread across `0`-`6`;
every other pair leans toward one extreme, matching the heatmap above.

## Takeaways

- `group_diff()`/`compute_group_diff()` return a full `g x g`
  matrix for any number of groups -- there is no special two-group code path.
- Groups with the same underlying effect show up directly as a small pairwise
  distance, without needing a separate equivalence test.
- `plot_multi_diff()`'s heatmap is a quick overview, not the final word: at
  enough features, OVL saturates there just as it does everywhere else --
  drill into `out$pairs[[...]]` (or use `group_diff()`'s own default) for the
  informative view.

## Related

*Basic pipeline* -- *Pooled vs. non-pooled fitting* -- *Analysis with a
block-diagonal partition* (for many features across many groups).
