---
title: "07 · Comparing groups with different sample sizes"
description: "Compare groups with different replicate counts and see how it affects posterior width."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{07 · Comparing groups with different sample sizes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

```{r setup, message = FALSE, warning = FALSE}
library(keRnel)
library(BayesOmics)
library(ggplot2)
```

One condition is cheap to sample, the other is rare or expensive -- the two
groups end up with very different replicate counts. This is fully
supported: a group's posterior scale is `n_obs + lambda_0`, using only
*that group's own* replicate count. The only real requirement is that both
groups share the same `Input` positions (so they share the same kernel
matrix and can be compared at all) -- nothing requires the same `N`.

## Simulate an unbalanced design

3 replicates for `Rare`, 15 for `Abundant`:

```{r}
set.seed(7)
kern_true <- variance_kernel(variance = 10) * se_kernel(length_scale = 80)
data <- simu_db_kernel(
  kernel        = kern_true,
  nb_id         = 6,
  nb_group      = 2,
  nb_sample     = c(3, 15),
  diff_group    = 0.8,
  var_sample    = 5,
  range_input   = c(0, 200),
  group_labels  = c("Rare", "Abundant"),
  integer_input = TRUE,
  input_grid    = TRUE
)
ggplot(data, aes(Input, Output, color = Group, group = interaction(Group, Sample))) +
  geom_line(alpha = 0.4) +
  geom_point(size = 1)
```

`Rare` has visibly fewer lines than `Abundant` -- the asymmetry the rest of
this example works through.

## Fit and compare, as usual

```{r}
kern <- variance_kernel(variance = 1) * se_kernel(length_scale = 50) +
  white_noise_kernel(noise = 1)
g1         <- data[data$Group == "Rare", ]
prior_mean <- mean(g1$Output)
hp_opt     <- fit_kernel(kern, g1, prior_mean = prior_mean, prior_cov = 1)
kern_opt   <- kupdate(kern,
                       variance     = hp_opt[["variance"]],
                       length_scale = hp_opt[["length_scale"]],
                       noise        = hp_opt[["noise"]])
posterior  <- posterior_mean(data, kern_opt, mu_0 = prior_mean, lambda_0 = 1)
```

Each group's own scale reflects its own sample size:

```{r}
c(Rare = posterior$groups$Rare$scale, Abundant = posterior$groups$Abundant$scale)
group_diff(posterior)
```

`Rare` has scale `4` (`3 + lambda_0`), `Abundant` has scale `16`
(`15 + lambda_0`) -- `Rare`'s posterior is proportionally wider, and the OVL
above already accounts for that difference correctly.

## Visualize the width difference

```{r}
samples <- sample_posterior(posterior, n = 2000)
plot_posterior_overlap(samples, group1 = "Rare", group2 = "Abundant",
                        id = unique(samples$ID)[1])
```

`Rare`'s density is visibly wider than `Abundant`'s -- fewer replicates,
less certainty about the mean -- even though both are centered where the
simulated effect places them.

## Takeaways

- Unequal `nb_sample` across groups needs no special handling: fit and
  compare exactly as in the balanced case.
- What *is* required is that both groups are measured at the same `Input`
  positions -- not the same replicate count.

## Related

[Basic pipeline](01_basic_pipeline.html) -- [Comparing more than two groups](05_multi_group.html).
