---
title: "08 · Using a kernel other than Squared Exponential"
description: "Beyond the Squared Exponential kernel: choosing and composing other kernels."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{08 · Using a kernel other than Squared Exponential}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 3.5)
```

```{r setup, message = FALSE, warning = FALSE}
library(keRnel)
library(BayesOmics)
library(ggplot2)
library(gridExtra)
```

The SE kernel assumes very smooth correlation. Nothing else in the pipeline
depends on that choice -- swapping the constructor is enough.

| Kernel                        | Use when                                      |
|--------------------------------|------------------------------------------------|
| `matern12_kernel()`            | Rough, non-smooth correlation                  |
| `matern32_kernel()`, `matern52_kernel()` | Intermediate smoothness, more realistic than SE for noisy biological signal |
| `periodic_kernel(length_scale, period)` | Cyclical structure (e.g. circadian, seasonal) |
| `linear_kernel()`               | A trend rather than a bump/dip                 |

Two worked examples below, each with the kernel's correlation-vs-distance
shape next to `keRnel::plot_kernel()`'s heatmap of the same kernel over the
actual feature positions -- the 1D curve on the left is exactly what the
2D banding on the right is built from.

`shape(kernel, d)` is the vectorized elementwise shape function every
stationary kernel provides -- `pairwise(kernel, x1, x2)` reduces to
`shape(kernel, distance_function(x1, x2))`, so this is the same correlation
a heatmap cell would show, just as a 1D curve over distance alone:

```{r}
kernel_shape_df <- function(kern, lag_max, label) {
  lag_seq <- seq(0, lag_max, length.out = 200)
  data.frame(lag = lag_seq, correlation = shape(kern, lag_seq), kernel = label)
}
```

## Matern 3/2: rougher than SE

```{r}
set.seed(6)
kern_true <- variance_kernel(variance = 10) * matern32_kernel(length_scale = 60)
data <- simu_db_kernel(kernel = kern_true, nb_id = 15, nb_group = 2, nb_sample = 8,
                        diff_group = 3, var_sample = 3, range_input = c(0, 200),
                        integer_input = TRUE, input_grid = TRUE)

p_shape <- ggplot(kernel_shape_df(matern32_kernel(length_scale = 60), 200, "Matern 3/2"),
                   aes(lag, correlation)) +
  geom_line(linewidth = 1, color = "#2C6E9E") +
  labs(title = "Kernel shape", x = "|Input_i - Input_j|", y = "correlation") +
  theme_minimal(base_size = 13)
p_heat <- plot_kernel(matern32_kernel(length_scale = 60), unique(data$Input))
grid.arrange(p_shape, p_heat, ncol = 2)
```

```{r}
kern <- variance_kernel(variance = 1) * matern32_kernel(length_scale = 50) +
  white_noise_kernel(noise = 1)
g1         <- data[data$Group == unique(data$Group)[1], ]
prior_mean <- mean(g1$Output)
hp_opt     <- fit_kernel(kern, g1, prior_mean = prior_mean, prior_cov = 1)
kern_opt   <- kupdate(kern, variance = hp_opt[["variance"]],
                       length_scale = hp_opt[["length_scale"]], noise = hp_opt[["noise"]])
posterior  <- posterior_mean(data, kern_opt, mu_0 = prior_mean, lambda_0 = 1)
group_diff(posterior)
```

## Periodic: cyclical structure

Same calls, `periodic_kernel(length_scale, period)` -- correlation now
oscillates with distance instead of decaying monotonically:

```{r}
set.seed(6)
kern_true <- variance_kernel(variance = 10) * periodic_kernel(length_scale = 15, period = 40)
data <- simu_db_kernel(kernel = kern_true, nb_id = 30, nb_group = 2, nb_sample = 8,
                        diff_group = 3, var_sample = 3, range_input = c(0, 200),
                        integer_input = TRUE, input_grid = TRUE)

p_shape <- ggplot(kernel_shape_df(periodic_kernel(length_scale = 15, period = 40), 200, "Periodic"),
                   aes(lag, correlation)) +
  geom_line(linewidth = 1, color = "#C9622F") +
  labs(title = "Kernel shape", x = "|Input_i - Input_j|", y = "correlation") +
  theme_minimal(base_size = 13)
p_heat <- plot_kernel(periodic_kernel(length_scale = 15, period = 40), unique(data$Input))
grid.arrange(p_shape, p_heat, ncol = 2)
```

```{r}
kern <- variance_kernel(variance = 1) * periodic_kernel(length_scale = 15, period = 40) +
  white_noise_kernel(noise = 1)
g1         <- data[data$Group == unique(data$Group)[1], ]
prior_mean <- mean(g1$Output)
hp_opt     <- fit_kernel(kern, g1, prior_mean = prior_mean, prior_cov = 1)
kern_opt   <- kupdate(kern, variance = hp_opt[["variance"]], length_scale = hp_opt[["length_scale"]],
                       period = hp_opt[["period"]], noise = hp_opt[["noise"]])
posterior  <- posterior_mean(data, kern_opt, mu_0 = prior_mean, lambda_0 = 1)
group_diff(posterior)
```

The heatmap's diagonal bands repeat every ~40 units, matching the kernel
shape's oscillation on the left -- an SE or Matern kernel could not produce
this pattern regardless of its length scale.

## Takeaways

- The kernel constructor is the only thing that changes between kernel
  choices -- `fit_kernel()`, `posterior_mean()` and every downstream metric
  are constructor-agnostic.
- The kernel shape plot is a direct preview of the correlation structure
  the fit will assume -- a monotonically decaying shape cannot capture
  periodic data, and vice versa.
- Every kernel here still benefits from a `+ white_noise_kernel(...)` term
  for the same numerical-stability reason as the SE case.

## Related

[Basic pipeline](01_basic_pipeline.html) -- [Input with more than one dimension](09_multi_dim_input.html).
