---
title: "Scaling up with a block-diagonal partition"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scaling up with a block-diagonal partition}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

```{r setup, message = FALSE, warning = FALSE}
library(keRnel)
library(BayesOmics)
```

`posterior_mean()` fits one kernel matrix covering every feature in a group
at once -- fine at 40 features, expensive well before a few thousand.
`fit_block_posterior()` splits the features into independent blocks first
and fits/inverts each block separately, trading a bit of cross-block
correlation for a much smaller matrix per fit.

## Partition and fit

`partition_by_range()` groups features whose `Input` falls in the same
interval -- nearby features stay correlated within a block, distant ones
are (correctly) treated as unrelated:

```{r}
set.seed(9)
kern_true <- variance_kernel(variance = 10) * se_kernel(length_scale = 15)
data <- simu_db_kernel(kernel = kern_true, nb_id = 40, nb_group = 2, nb_sample = 8,
                        diff_group = 0.8, var_sample = 5, range_input = c(0, 200),
                        integer_input = TRUE, input_grid = TRUE)

kern_template <- variance_kernel(variance = 1) * se_kernel(length_scale = 50) +
  white_noise_kernel(noise = 1)
partition <- partition_by_range(n_bins = 4)
fit <- fit_block_posterior(data, partition, kern = kern_template, pooled = TRUE,
                            prior_mean = mean(data$Output), prior_cov = 1)
fit$blocks
```

`kern` is an *unfitted* template here: `fit_kernel()` is run once per block
rather than once for the whole dataset.

## Each block is a normal posterior_mean() result

`fit$block_results[[b]]` is exactly what `posterior_mean()` would have
returned on that block alone -- every existing function (`group_diff()`,
`compute_group_diff()`, ...) works on it unchanged:

```{r}
sapply(fit$blocks, function(b) group_diff(fit$block_results[[b]])[1, 2])
```

## Reassembling a full-length result

`get_muk_block()`/`get_sigmak_block()` concatenate the per-block pieces back
into one posterior mean vector / list of block covariances for a group,
when you need the full picture rather than block-by-block:

```{r}
muk <- get_muk_block(fit, unique(data$Group)[1])
length(muk)
sapply(get_sigmak_block(fit, unique(data$Group)[1]), dim)
```

Each block is 10x10 here, never the full 40x40.

Reassembling the blocks side by side also makes the trade-off visible: the
full covariance is block-diagonal, every off-block entry exactly `0` --
cross-block correlation is what's being given up in exchange for a much
smaller matrix per fit:

```{r, fig.alt = "Heatmap of the reassembled 40x40 covariance matrix, showing four dense diagonal blocks and zero everywhere off-block"}
sigma_blocks <- get_sigmak_block(fit, unique(data$Group)[1])
full_dim  <- sum(sapply(sigma_blocks, nrow))
full_mat  <- matrix(0, full_dim, full_dim)
ids_order <- unlist(lapply(sigma_blocks, rownames))
offset <- 0
for (b in sigma_blocks) {
  n <- nrow(b)
  full_mat[(offset + 1):(offset + n), (offset + 1):(offset + n)] <- b
  offset <- offset + n
}
rownames(full_mat) <- ids_order
colnames(full_mat) <- ids_order
mat_df <- as.data.frame(as.table(full_mat))
names(mat_df) <- c("Row", "Col", "Value")
mat_df$Row <- factor(mat_df$Row, levels = ids_order)
mat_df$Col <- factor(mat_df$Col, levels = rev(ids_order))
ggplot2::ggplot(mat_df, ggplot2::aes(Row, Col, fill = Value)) +
  ggplot2::geom_tile() +
  ggplot2::scale_fill_gradient(low = "white", high = "#5E72A4") +
  ggplot2::labs(title = "Reassembled covariance: block-diagonal structure", x = NULL, y = NULL) +
  ggplot2::theme(axis.text = ggplot2::element_blank(), axis.ticks = ggplot2::element_blank())
```

## Takeaways

- Passing an unfitted kernel *template* (not `NULL`) fits it independently
  per block via `fit_kernel()` -- the same object you would use directly with
  `posterior_mean()`.
- Each block behaves as a standalone `posterior_mean()` result: no new
  metric API to learn, just call the usual functions per block.

## Related

*Basic pipeline* -- *Comparing more than two groups*.
