---
title: "Working from real data instead of a simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working from real data instead of a simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

```{r setup, message = FALSE, warning = FALSE}
library(keRnel)
library(BayesOmics)
library(ggplot2)
```

Every earlier example started from `simu_db_kernel()`. Real data needs the
same five columns, but getting there involves choices `simu_db_kernel()`
made for you: which axis is `Input`, what counts as a replicate, and
whether every group actually shares the same `ID`s. `datasets::ChickWeight`
(shipped with base R, no omics data needed to illustrate this) stands in
for a real longitudinal measurement:

| BayesOmics concept | Here                                   |
|---------------------|-----------------------------------------|
| `Input`             | age in days                             |
| `ID`                | one measurement day (`"Day_0"`, ...)    |
| `Sample`             | one chick (a real biological replicate) |
| `Group`             | diet                                    |
| `Output`            | body weight (g)                         |

## Restrict to a shared `ID` set

`group_diff()`/`compute_group_diff()`/`calculate_group_overlaps()` require both groups to
share the exact same set of `ID`s. Real longitudinal data has dropout, so
this has to be enforced explicitly -- chicks missing any of the 11 standard
weighing days are dropped:

```{r}
data("ChickWeight", package = "datasets")
cw <- as.data.frame(ChickWeight)
cw$Diet  <- as.character(cw$Diet)
cw$Chick <- as.character(cw$Chick)

std_days     <- c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18, 20)
has_all_days <- function(sub) all(std_days %in% sub$Time)
ok_chicks    <- names(which(vapply(split(cw, cw$Chick), has_all_days, logical(1))))
cw2          <- cw[cw$Chick %in% ok_chicks & cw$Time %in% std_days, ]
length(ok_chicks)
```

## Reshape into the long format

```{r}
build_group <- function(diet_label, group_label) {
  sub    <- cw2[cw2$Diet == diet_label, ]
  chicks <- sort(unique(sub$Chick))
  do.call(rbind, lapply(seq_along(chicks), function(s) {
    one <- sub[sub$Chick == chicks[s], ]
    one <- one[order(one$Time), ]
    data.frame(ID = paste0("Day_", one$Time), Group = group_label, Sample = s,
               Input = one$Time, Output = one$weight, stringsAsFactors = FALSE)
  }))
}
data_all <- rbind(build_group("1", "Diet1"), build_group("3", "Diet3"))
head(data_all, 4)
```

Each chick is one replicate line, colored by diet -- the same long-format
data frame every other article's `simu_db_kernel()` output has, just built
by hand here instead of simulated:

```{r, fig.alt = "Individual chick growth curves (weight vs. age in days), one line per chick, coloured by diet"}
ggplot(data_all, aes(Input, Output, color = Group, group = interaction(Group, Sample))) +
  geom_line(alpha = 0.4) +
  geom_point(size = 1) +
  labs(x = "Age (days)", y = "Weight (g)")
```

## Fit and compare, as usual

Nothing pipeline-specific changes from here -- `prior_cov` is set from the
data's own inter-chick variance rather than guessed:

```{r}
prior_cov_val <- mean(with(data_all, tapply(Output, list(Group, Input), var)), na.rm = TRUE)
kern <- variance_kernel(variance = 2000) * se_kernel(length_scale = 5) +
  white_noise_kernel(noise = 100)
d1     <- data_all[data_all$Group == "Diet1", ]
hp_opt <- fit_kernel(kern, d1, prior_mean = mean(d1$Output), prior_cov = prior_cov_val)
kern_opt  <- kupdate(kern, variance = hp_opt[["variance"]],
                      length_scale = hp_opt[["length_scale"]], noise = hp_opt[["noise"]])
posterior <- posterior_mean(data_all, kern_opt, mu_0 = mean(data_all$Output), lambda_0 = 1)
```

```{r}
calculate_group_overlaps(posterior)
compute_group_diff(posterior, per_feature_metric(wasserstein_metric(), power = 0.5))
```

The joint OVL is a numerical `0` -- exactly the saturation described in the
metric-choice article, expected once two full growth curves are compared
across 11 days at once. The per-feature Wasserstein (~38 g) stays
interpretable: on average, the two diets differ by about 38 g at a given day.
This is exactly the case `group_diff(posterior)` (no metric specified) would
have picked automatically: with 11 features, above its default threshold,
it returns the per-feature Wasserstein directly rather than the saturated OVL.

## Takeaways

- Real data needs an explicit ID-alignment step that a simulator gives you
  for free -- decide it before fitting, not after an error.
- Choosing which real axis maps to `ID` vs. `Sample` vs. `Input` is a
  domain decision `simu_db_kernel()` cannot make for you.

## Bonus: let an AI assistant write the reshaping code for you

The two steps above -- restricting to a shared `ID` set, then reshaping to
the long format -- are mechanical once you know which of your raw columns
plays which role, but the mapping itself is dataset-specific. The following
prompt (fill in the bracketed description of your own data) gives an AI
assistant everything it needs to write that reshaping code directly,
following the same rules `ChickWeight` was just reshaped by:

```text
I have a dataset I want to reshape into the long format required by the R
package BayesOmics (Bayesian kernel-based differential analysis). BayesOmics
needs a data frame with these columns:

- ID: the feature identifier (e.g. a genomic position, a protein, a
  measurement day/timepoint) -- the axis along which correlation is modeled.
- Group: the experimental condition/group each observation belongs to.
- Sample: the replicate identifier within a group (e.g. a subject, a
  biological replicate, an individual).
- Input: the NUMERIC covariate the kernel operates on (often the same
  information as ID, but as a number -- e.g. genomic position, age, dose).
  Within a single Group, each ID must map to exactly one Input value.
- Output: the observed numeric measurement.
- (Optional) Input_ID: only needed if Input has more than one dimension per
  feature (e.g. two covariates at once) -- one row per Input_ID, with Output
  repeated identically across the rows of one observation.

Constraints to respect:
1. Exactly one row per (ID, Group, Sample[, Input_ID]).
2. Within a Group, each ID must map to exactly one Input value (no
   duplicated/inconsistent Input for the same ID).
3. Every Group I want to compare should share the exact same set of IDs --
   drop or filter out whatever breaks this (e.g. missing data/dropout),
   and report how much was dropped.
4. Group and ID should be character columns, not factors.

Here is a description of my raw dataset: [describe your raw column names and
what each one means -- which column is the group/condition, which is the
replicate, which is the numeric position/covariate, which is the
measurement, and any known data-quality issues like missing values].

Please write R code (base R or dplyr/tidyr, whichever is cleaner) that:
1. Loads/reads my raw data.
2. Identifies and handles any IDs missing from at least one group (report
   how many rows/subjects are dropped and why).
3. Reshapes it into the BayesOmics long format described above.
4. Prints head() of the result, and confirms for at least two groups that
   they share the exact same set of ID values (e.g. via setequal()).
```

## Related

*Choosing a metric: OVL vs. Wasserstein* -- *Comparing groups with different
sample sizes*.
