---
title: "11 · Performance at a glance"
description: "Timings versus feature count, and when to switch to block-diagonal fitting."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{11 · Performance at a glance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The cost of an analysis is driven by the number of features `d` in a group:
fitting the kernel repeatedly factorises a `d x d` covariance matrix, which
grows roughly with `d^3`. Measured once on a laptop, with two groups of eight
replicates, a Squared Exponential plus noise kernel, and default optimiser
settings:

| Features `d` | `fit_kernel()` (one group) | `posterior_mean()` | Monolithic total | `fit_block_posterior()` total |
|---:|---:|---:|---:|---:|
| 25 | 0.23 s | 0.12 s | 0.35 s | 0.58 s |
| 50 | 0.28 s | 0.13 s | 0.41 s | 0.94 s |
| 100 | 0.64 s | 0.19 s | 0.83 s | 1.53 s |
| 200 | 0.89 s | 0.28 s | 1.2 s | 2.7 s |
| 400 | 3.7 s | 0.61 s | 4.3 s | 2.6 s |
| 800 | 25.6 s | 1.1 s | 26.7 s | 4.8 s |
| 1200 | 100.2 s | 2.3 s | 102.5 s | 9.0 s |

Blocks hold about 25 features up to `d = 200` and about 100 beyond.

## What to take from it

- **Up to a few hundred features, stay monolithic.** Everything finishes in
  seconds, and `posterior_mean()` keeps every cross-feature correlation.
  Block-diagonal fitting is even slower here, because it pays for one fit per
  block.
- **The kernel fit is the bottleneck, not the posterior.** From `d = 400`
  the fit dominates; `posterior_mean()` stays cheap.
- **Past roughly 300 to 400 features, split into blocks.** The gain grows
  quickly (about 11x at `d = 1200`), at the price of ignoring correlation
  between blocks. See [Block-diagonal partition](10_block_diagonal.html).

## Reproduce it on your machine

Timings depend on your hardware, so measure your own case:

```{r, eval = FALSE}
library(keRnel)
library(BayesOmics)

d <- 400
kern_true <- variance_kernel(variance = 10) * se_kernel(length_scale = 15)
data <- simu_db_kernel(kernel = kern_true, nb_id = d, nb_group = 2, nb_sample = 8,
                       diff_group = 0.8, var_sample = 5, range_input = c(0, 4 * d),
                       integer_input = TRUE, input_grid = TRUE)
kern <- variance_kernel(variance = 1) * se_kernel(length_scale = 50) +
  white_noise_kernel(noise = 1)
prior_mean <- mean(data$Output)

system.time(
  fit_kernel(kern, data[data$Group == unique(data$Group)[1], ],
             prior_mean = prior_mean, prior_cov = 1)
)
system.time(
  fit_block_posterior(data, partition_by_range(n_bins = d %/% 100), kern = kern,
                      pooled = TRUE, prior_mean = prior_mean, prior_cov = 1)
)
```

## Related

[Basic pipeline](01_basic_pipeline.html) --
[Scaling up with a block-diagonal partition](10_block_diagonal.html).
