---
title: "Get started with BayesOmics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with BayesOmics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup, message=FALSE, warning=FALSE}
library(keRnel)
library(BayesOmics)
library(ggplot2)
```

The BayesOmics package provides a suite of tools for performing Bayesian kernel-based differential analysis.

## Data format

Every function in BayesOmics expects a long-format data frame with these
columns:

| Column     | Meaning                                                                    |
|------------|-----------------------------------------------------------------------------|
| `ID`       | Feature identifier (e.g. a peptide, a CpG site, a metabolite)              |
| `Group`    | Experimental condition the observation belongs to                          |
| `Sample`   | Replicate identifier within a group                                        |
| `Input_ID` | Which Input dimension a row belongs to (see below)                        |
| `Input`    | The covariate the kernel operates on (e.g. genomic position)              |
| `Output`   | The observed measurement                                                   |

A single-covariate analysis (the case throughout this vignette) has one
`Input_ID` value per feature; a data frame with just `Input` (no `Input_ID`)
also works unchanged -- BayesOmics fills in `Input_ID` automatically. A
multi-dimensional Input (several covariates per feature, e.g. genomic
position *and* measurement time) instead has one row per `Input_ID` value,
with `Output` repeated identically across the rows of one observation.

## Choosing and initializing a kernel

The kernel encodes which features should be treated as correlated. The most
common choice is the **Squared Exponential (SE) kernel**:

$$
K_{\text{SE}}(x, x') = \sigma^2 \exp\!\left(-\frac{\|x - x'\|^2}{2\ell^2}\right)
$$

where $\sigma^2$ is the signal variance and $\ell$ is the length scale  -  how
far apart two features can be before their correlation becomes negligible.

Create an SE kernel and set its hyperparameters to the values that will be
used to generate synthetic data:

```{r}
kern_true <- variance_kernel(variance = 10) * se_kernel(length_scale = 80)
```

## Simulating data

`simu_db_kernel()` generates a dataset whose cross-feature covariance matches
the generative model assumed by `multi_posterior_mean()`: for each group,
the vector of outputs across all features is drawn jointly from a
multivariate normal whose covariance is given by the kernel, plus independent
measurement noise. Each replicate is an independent draw from that same
distribution:

$$
y_n \mid \mu_g \sim \mathcal{N}(\mu_g,\; \Sigma_\theta + \sigma^2 I),
$$

```{r}
set.seed(42)
data <- simu_db_kernel(
  kernel      = kern_true,
  nb_id       = 10,
  nb_group    = 2,
  nb_sample   = 8,
  diff_group  = 0.8,
  var_sample  = 5,
  range_input = c(0, 200)
)
head(data)
```


## Combining kernels: adding a Noise term

Kernels can be combined with `+` (additive) or `*` (product). Adding a
`NoiseKernel` nugget to absorb feature-level measurement noise independently
of the spatial signal is not just a nice-to-have here: with realistic sample
sizes, fitting a spatial kernel *alone* to noisy data tends to produce an
almost-singular covariance matrix (see the note further down), which silently
breaks the differential-analysis step. Always fit and reuse the combined
kernel, not just its spatial part:

```{r}
se_part    <- variance_kernel(variance = 1) * se_kernel(length_scale = 1)
noise_part <- white_noise_kernel(noise = 1)
kern       <- se_part + noise_part
```

Using an SE + Noise kernel for optimization separates the spatial signal
($\sigma^2$, $\ell$) from the measurement noise ($c$).

## Optimizing hyperparameters

`optim_hp()` fits all hyperparameters by maximizing the marginal likelihood of
the output values.

```{r}
groupe1=data[data$Group == unique(data$Group)[1], ]
prior_mean   <- mean(groupe1$Output)

hp_opt <- optim_hp(kern, groupe1, prior_mean = prior_mean, prior_cov = 1)
hp_opt
```

`length_scale` (~ 98) lands in the same ballpark as its simulation value
(80), confirming that correlation extends over most of the 200-unit axis.
`variance` (~ 15.5) and `noise` (~ 4.5) partially trade off against each
other with only 8 replicates per group -- their individual values are less
tightly recovered than `length_scale` -- but that is not a problem: what
matters for the next step is fitting *both* together, so the resulting
covariance correctly separates spatial signal from measurement noise.

Update the kernel with the fitted values, **keeping both the spatial and noise hyperparameters**:

```{r}
kern_opt <- do.call(kupdate, c(list(kern), as.list(hp_opt)))
```

Do not rebuild a plain SE kernel from only `variance` and `length_scale`
here, discarding `noise`. With this many features and a length scale this
large relative to the input range, a pure-SE kernel matrix is nearly
singular (its condition number can exceed 1e9). The differential analysis
below still runs on such a matrix, but the resulting overlap coefficient
becomes numerically meaningless -- typically collapsing to exactly 0
regardless of the true difference between groups. Keeping the fitted
`noise` term is what keeps the covariance well-conditioned.

## Computing posteriors

`multi_posterior_mean()` applies the Normal-Normal conjugate update to every
group at once:

$$
p(\boldsymbol{\mu} \mid y_1, \dots, y_N,\, \Sigma_{\hat\theta})
= \mathcal{N}\!\left(
    \boldsymbol{\mu};\;
    \frac{\lambda_0\,\mu_0 + \sum_{n=1}^N y_n}{N + \lambda_0},\;
    \frac{1}{N + \lambda_0}\,\Sigma_{\hat\theta}
  \right)
$$

where $\Sigma_{\hat\theta}$ is the kernel matrix at the optimized
hyperparameters, $\mu_0$ and $\lambda_0$ are the prior mean and precision, and
$N$ is the number of replicates.

```{r}
posterior <- multi_posterior_mean(data, kern_opt, mu_0 = prior_mean, lambda_0 = 1)
print(posterior)
```

Draw samples from the posteriors for plotting and the overlap calculation:

```{r}
samples <- sample_posterior(posterior, n = 2000)
```

## Differential analysis

`calculate_group_overlaps()` returns the pairwise Overlapping Coefficient (OVL)
between every pair of groups. Groups with different sample sizes are fully
supported  -  see `vignette("sample_size_scenarios")` for a detailed exploration
of how replicate counts affect posterior width and the resulting OVL.
OVL = 1 means identical posteriors (no
differential signal); OVL ~ 0 means fully separated profiles (strong
differential signal):

$$
\text{OVL} = 2\,\Phi\!\left(-\tfrac{\delta}{2}\right)
$$

where $\delta$ is the Mahalanobis distance between the two groups' posterior
means under the posterior covariance of the reference group, and $\Phi$ is the
standard normal CDF.

```{r}
calculate_group_overlaps(posterior)
```

Overlapping coefficient can be seen as a single-number summary of the differential signal on the whole region of interest. It is useful for ranking features, but it is sensible to curse of dimensionality: the more features you include, the more likely it is that the joint overlap will be low. The OVL is a global measure, and it does not tell you which features are driving the difference.
`compute_multi_diff()` complements the single OVL number with an
uncertainty-aware breakdown: from the posterior draws, it computes the
empirical distribution of the number of features for which group 1's draw
exceeds group 2's, and (when given `results`) attaches the same OVL computed
above for reference. `plot_multi_diff()` turns this into a compact figure --
one panel per group pair, titled with that pair's OVL, plus the
posterior-mean panel already seen above:

```{r}
multi_diff <- compute_multi_diff(samples, results = posterior)
multi_diff$Diff_proba
```

```{r, fig.alt = "Empirical distribution of the number of features where group 1's posterior draw exceeds group 2's, spread broadly across 0 to 10 features rather than spiked at either extreme, alongside the posterior-mean-per-feature panel"}
plot_multi_diff(multi_diff)
```

The bar-chart panel is spread broadly across 0 to 10 features, rather than
spiked at either extreme -- group 1 does not exceed group 2 on every feature,
nor on none. This is consistent with a real, but partial and non-caricatural,
differential signal: exactly the kind of nuance the single OVL number above
cannot show on its own.

## Individual feature plots
`plot_posterior_mean()` shows the posterior mean at every feature, coloured by
group  -  useful to see which features drive the differential signal:

```{r}
plot_posterior_mean(samples)
```

`plot_posterior_overlap()` overlays both groups' posterior densities for one
feature and shades the region of overlap, giving a visual counterpart of the
OVL coefficient:

```{r}
plot_posterior_overlap(samples,
                       group1 = unique(samples$Group)[1],
                       group2 = unique(samples$Group)[2],
                       id     = unique(samples$ID)[1])
```

## Go further

| Vignette | What you will find |
|---|---|
| `vignette("omics-analysis")` | Full worked examples  -  realistic biological context, kernel composition (SE + Noise), HP interpretation, two-group and four-group dose-response designs, and every plot function. |
| `vignette("sample_size_scenarios")` | How unbalanced designs affect posterior width and the OVL, and why the comparison remains valid with strongly unequal sample sizes. |
| `vignette("troubleshooting")` | Exact error messages you may encounter, what causes each of them, and the targeted fix. |
