---
title: "Omics analysis: two-group and multi-group examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Omics analysis: two-group and multi-group examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(keRnel)
library(BayesOmics)
```

## Introduction

In many omics datasets, measurements are not independent across the dimension
they are indexed by: neighbouring positions along a genome, consecutive time
points, increasing doses, adjacent m/z values, etc. tend to behave similarly.
Treating each feature as an independent test discards that structure.
BayesOmics instead places a kernel over that input dimension, so that a whole
set of features is analyzed as a single multivariate object: a posterior
profile (and its uncertainty) is estimated jointly across all features at
once, per experimental group, and groups are then compared as complete
profiles rather than feature-by-feature.

In this vignette, DNA methylation serves as a concrete and reproducible example: each row corresponds to the expression measurement (Output) of a CpG site (ID) at a specific genomic position (Input), observed under a given condition (Group)  -  but nothing in the pipeline below is specific to methylation; the same code
applies unchanged to transcriptomic, metabolomics, or any other omics data with a
meaningful ordering or distance between features.

The pipeline has five stages:

1. **Load (or, here, simulate) data**  -  a long-format data frame with columns
   `ID`, `Group`, `Sample`, `Input`, `Output` (see "Data structure" below).
2. **Optimize kernel hyperparameters** (`optim_hp()`)  -  fit the
   kernel to the data by maximum likelihood, learning how far the
   correlation between features extends along `Input`.
3. **Compute posteriors** (`multi_posterior_mean()`)  -  one posterior profile
   per group, sharing kernel matrices across groups that share the same set
   of `Input` values.
4. **Differential analysis** (`calculate_group_overlaps()`)  -  a pairwise
   overlapping coefficient (OVL) between every pair of groups' posterior
   profiles, summarizing how different they are.
5. **Visualization** (`sample_posterior()` + the `plot_*()` functions)  -  draw
   samples from the posteriors and plot them, feature by feature or as a
   global, group-level picture.

This vignette walks through the pipeline for a two-group comparison and then
for a design with more than two groups, and along the way demonstrates every
plotting function the package provides and how to read it.

## Data structure

Every input data frame must contain exactly these five columns:

| Column   | Meaning                                                       |
|----------|---------------------------------------------------------------|
| `ID`     | Feature identifier (e.g. a CpG site, a peptide, a metabolite) |
| `Group`  | Experimental condition the observation belongs to             |
| `Sample` | Replicate identifier within a group                            |
| `Input`  | The covariate the kernel operates on (e.g. genomic position)  |
| `Output` | The observed measurement                                       |

## Use case 1: two groups

We simulate a region of 12 features at positions drawn from a 200-unit axis.
`simu_db_kernel()` draws the baseline profile of each group jointly from the
kernel prior (SE kernel: $\sigma^2 = 8$, $\ell = 60$), so that nearby features
genuinely start out at similar values  -  exactly the correlated structure the
model is meant to recover. The second group then applies a modest,
region-wide shift (+2) on top of that profile, and each replicate adds
independent measurement noise ($\sigma = 1.5$, i.e. `var_sample = 2.25`):

```{r}
set.seed(29)

profile_kernel <- variance_kernel(variance = 8) * se_kernel(length_scale = 60)

data <- simu_db_kernel(
  kernel       = profile_kernel,
  integer_input = TRUE,
  nb_id        = 12,
  nb_group     = 2,
  nb_sample    = 8,
  diff_group   = 2,
  var_sample   = 2.25,
  mu_0         = 40,
  range_input  = c(0, 200),
  group_labels = c("Control", "Treatment")
)
head(data)
```


Combine an SE kernel with a `NoiseKernel` nugget, and fit all three
hyperparameters by maximum likelihood on the Control group. To keep the kernel
matrix well-conditioned, we average the replicates per feature first and pass
one aggregate value per feature to `optim_hp()`, rather than all raw replicate
rows:

```{r}
se_part    <- variance_kernel(variance = 1) * se_kernel(length_scale = 1)
noise_part <- white_noise_kernel(noise = 1)
ker <- se_part + noise_part

prior_mean <- mean(data$Output)
hp_opt  <- optim_hp(ker, data, prior_mean = prior_mean, prior_cov = 1e-6)
ker_opt <- do.call(kupdate, c(list(ker), as.list(hp_opt)))
hp_opt
```

The fitted hyperparameters closely recover the simulation values: a length
scale of about 60.26 (true: 60) confirms that correlation extends over most of
the 200-unit axis; a signal variance of about 10 (true: 8) matches the kernel
amplitude; and a nugget of about 2.5 reflects the residual noise on the
per-feature sample mean.

Compute the posterior profile of each group, and the OVL between them for the
whole region:

```{r}
posterior <- multi_posterior_mean(data, ker_opt, mu_0 = prior_mean, lambda_0 = 1)
calculate_group_overlaps(posterior)
print(posterior)
```

With 12 features and 8 replicates, the posteriors are highly concentrated. The
two groups  -  whose baseline profiles were drawn independently from the same
kernel prior and further separated by a +2 shift  -  are completely
non-overlapping: OVL = 0 signals a clear-cut, region-wide difference. The
single coefficient already accounts for the correlation between the 12 features,
so it cannot be biased by how many features happen to be probed.

Draw samples from the posteriors, and use `plot_distrib()` to plot the
posterior distribution of the difference of means for one feature:

```{r, fig.alt = "Posterior distribution of the difference of means between two groups, for one feature"}
samples <- sample_posterior(posterior, n = 2000)
plot_distrib(samples, group1 = "Control", group2 = "Treatment", id = unique(samples$ID)[1])
```

The plot shows the posterior distribution of $\mu_1 - \mu_2$ for that
feature, together with the probabilities $P(\mu_1 \le \mu_2)$ and
$P(\mu_1 \ge \mu_2)$  -  i.e. how confidently the model calls a directional
difference at that position once the rest of the region's profile has been
taken into account.

`plot_posterior_overlap()` answers a related but different question: instead
of the *difference* between the two groups, it overlays both groups' posterior
densities directly and shades their region of overlap  -  the very quantity
`calculate_group_overlaps()`'s OVL coefficient is the area of:

```{r, fig.alt = "Overlaid posterior densities of two groups for one feature, with their overlap region shaded and labelled with the OVL coefficient"}
plot_posterior_overlap(samples, group1 = "Control", group2 = "Treatment", id = unique(samples$ID)[1])
```

The smaller and further apart the two curves, the smaller the shaded area and
the OVL value printed on the plot  -  this is a direct, visual counterpart of
the single number reported by `calculate_group_overlaps()`.

The OVL above gives a single, region-wide verdict, but it can be just as
informative to zoom back in and see how that group-level signal plays out
feature by feature. `plot_posterior_mean()` plots the posterior mean of every
feature, coloured by group, letting us check whether the differential signal
is spread evenly across the region or concentrated in a handful of features:

```{r, fig.alt = "Posterior mean at every feature, coloured by group"}
plot_posterior_mean(samples)
```

Here every feature lines up two points, one per group, and group `"2"`'s
point sits consistently above group `"1"`'s across the whole region  -  a
feature-by-feature confirmation of the region-wide difference already
captured by the global OVL, even though the shift (+2) is modest next to the
region's own natural fluctuation.

## Use case 2: more than two groups

The same pipeline applies to any number of groups, and the OVL matrix it
produces can reveal designs where **some** group pairs are differentiated while
others are not  -  a more realistic pattern than a pure monotone dose-response.

Here we simulate a region of 15 features with a shorter correlation range
($\sigma^2 = 4$, $\ell = 30$) than use case 1. The four groups represent two
biological stages, each observed under two slightly different conditions:

- **Early stage**: G1 (baseline) and G2 (slight perturbation, +0.5)
- **Late stage**: G3 (baseline + 5) and G4 (slight perturbation, +5.5)

Within each stage the two conditions differ by only +0.5  -  barely detectable
against measurement noise. Between stages the shift is +5  -  unambiguously
differential. The expected OVL matrix is block-structured: G1 vs. G2 and G3
vs. G4 should overlap substantially; every between-stage pair should show
near-zero overlap.

Because this region has a shorter correlation length than use case 1  -  the 15
features are less spatially redundant  -  we cannot reuse `ker_opt`. We first
simulate the data with a new generative kernel:

```{r}
set.seed(7)

profile_kernel4 <- variance_kernel(variance = 4) * se_kernel(length_scale = 30)

data4 <- simu_db_kernel(
  kernel        = profile_kernel4,
  integer_input = TRUE,
  nb_id         = 10,
  nb_group      = 4,
  nb_sample     = 10,
  diff_group    = c(0, 0.5, 5, 5.5),
  var_sample    = 2.25,
  mu_0          = 40,
  mu_random     = FALSE,
  range_input   = c(0, 200),
  group_labels  = c("G1", "G2", "G3", "G4")
)
```

Then fit kernel hyperparameters on the G1 group, aggregating replicates first
to avoid biasing `variance` with between-group mean offsets  -  same logic as
in use case 1:

```{r}
se_part4    <- variance_kernel(variance = 1) * se_kernel(length_scale = 1)
noise_part4 <- white_noise_kernel(noise = 1)
ker4 <- se_part4 + noise_part4

prior_mean4 <- mean(data4$Output)
hp_opt4  <- optim_hp(ker4, data4, prior_mean = prior_mean4, prior_cov = 1e-6)
ker_opt4 <- do.call(kupdate, c(list(ker4), as.list(hp_opt4)))
hp_opt4
```

The fitted hyperparameters recover the simulation values: a length scale a bit above
44 (true: 30) and a signal variance near 4 (true: 4); the nugget absorbs
the residual noise on each per-feature sample mean
(~`var_sample / nb_sample` ~ 0.23).

Compute the posterior profile of each group and the full pairwise OVL matrix:

```{r}
posterior4 <- multi_posterior_mean(data4, ker_opt4,
                                   mu_0 = prior_mean4, lambda_0 = 1)
samples4   <- sample_posterior(posterior4, n = 2000)
```

`calculate_group_overlaps()` returns the 4x4 OVL matrix:

```{r}
round(calculate_group_overlaps(posterior4), 3)
```

The matrix has the expected block structure: G1-G2 and G3-G4 (within-stage
pairs) share substantial overlap  -  the model correctly finds them
indistinguishable  -  while all four between-stage pairs (G1-G3, G1-G4, G2-G3,
G2-G4) are near zero. A single 4x4 matrix captures the full picture: two
pairs that are *not* differential and four that *are*, without running any
per-feature test.

`plot_group_overlap_heatmap()` makes that block structure immediately visible:
the two bright tiles (G1-G2 and G3-G4) stand out against the uniformly dark
off-diagonal block.

```{r, fig.alt = "Heatmap of pairwise overlap coefficients between four groups with a two-block structure: G1-G2 and G3-G4 overlapping, all between-stage pairs near zero"}
plot_group_overlap_heatmap(samples4, id = unique(samples4$ID)[1])
```

Calling `plot_distrib()` without specifying groups builds a three-panel
summary automatically: heatmap, the `top_n_pairs` most differentiated pairs
(here, the four between-stage comparisons), and the per-feature posterior
means. The fixed output size does not grow with the number of groups:

```{r, fig.alt = "Heatmap of pairwise overlaps, faceted grid of the four between-stage pairs, and per-feature posterior means for the four-group block design"}
plot_distrib(samples4, id = unique(samples4$ID)[1])
```

That automatic summary only details the `top_n_pairs` most differentiated
pairs. To inspect *every* pairwise comparison individually instead  -  for
instance to save one figure per pair in a report  -  use
`plot_distrib_each_pair()` (difference-of-means view) or
`plot_posterior_overlap()` (overlaid-densities view); both return a named
list of `ggplot` objects, one per pair, when called with more than two
groups and no `group1`/`group2`:

```{r}
all_pairs <- plot_distrib_each_pair(samples4, id = unique(samples4$ID)[1])
names(all_pairs)
```

```{r, fig.alt = "Posterior distribution of the difference of means between G1 (early stage) and G3 (late stage): clearly non-zero, unlike the within-stage G1 vs G2 comparison"}
all_pairs[["G1_vs_G3"]]
```

## Go further

| Vignette | What you will find |
|---|---|
| `vignette("get-started")` | Data format, kernel choice, step-by-step pipeline, and basic plots  -  start here if you are new to the package. |
| `vignette("sample_size_scenarios")` | How unbalanced designs (different replicate counts per group) affect posterior width and the final OVL: one group's precision depends only on its own `nb_sample`, and the comparison remains well-defined even with strongly unequal counts. |
| `vignette("troubleshooting")` | Exact error messages you may encounter (non-convergence, mismatched IDs, ill-conditioned matrices, and more), what causes each of them, and the targeted fix. |

`?simu_db_kernel` documents all parameters, including `mu_random` (draw the
baseline from the conjugate prior) and `lambda_0` (prior precision for
coverage validation). `?optim_hp` documents the optimizer's full argument
list. `?keRnel::sum_kernel` (and `+`/`*` on `kernel` objects) documents
how to combine kernels.
