---
title: "Unbalanced Designs: Differential Analysis with Different Sample Sizes per Group"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Unbalanced Designs: Differential Analysis with Different Sample Sizes per Group}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(keRnel)
devtools::load_all(".")
library(ggplot2)
```

## Groups don't need the same number of replicates

A common real-world constraint: one condition is cheap/abundant to sample, the other is
expensive or rare, so the two groups end up with very different replicate counts. BayesOmics
handles this natively. The posterior for a single group is

\[
  p(\mu \mid y_1,\dots,y_N,\Sigma_{\hat\theta}) = \mathcal{N}\!\left(\mu;\ \dfrac{\lambda_0\mu_0+\sum_n y_n}{N+\lambda_0},\ \dfrac{1}{N+\lambda_0}\Sigma_{\hat\theta}\right)
\]

`N` (`nb_sample`) only ever appears through *that group's own* count -- nothing here requires
the two groups being compared to share the same `N`. The **only** real requirement is that
both groups are measured at the **same `Input` values**, so that they share the same kernel
matrix \(\Sigma_{\hat\theta}\) (same `kernel_key` internally) and can be compared at all.
That is also exactly what makes `calculate_group_overlaps()`'s closed-form overlap work when
the two groups' scales differ (\(\mathrm{scale}_1 = N_1+\lambda_0 \ne \mathrm{scale}_2 =
N_2+\lambda_0\)): same \(\Sigma_{\hat\theta}\), different scaling.

This vignette simulates one fixed, spatially-correlated ground truth (same `Input` positions
for every group) and varies **how many replicates each group gets, independently of the
other** -- including strongly *unbalanced* designs -- to show that the differential analysis
still works, and to visualize exactly what changes.

## A fixed ground truth, several (Control, Treatment) sample-size pairs

As in the introductory vignette, the *true* methylation profile is itself spatially
correlated (drawn from an SE kernel over genomic position), and Treatment shifts that
profile by a fixed amount. We simulate this ground truth **once**  -  a base dataset
with 30 replicates per group, the maximum used in any scenario below  -  using
`simu_db_kernel()` with `mu_random = TRUE` so that each group draws a genuinely
spatially-varying baseline. Each scenario is then a subset of that base dataset:
only the number of retained replicates changes, and it changes independently per group:

```{r}
set.seed(42)

profile_kernel <- variance_kernel(variance = 6) * se_kernel(length_scale = 25)

data_base <- simu_db_kernel(
  kernel       = profile_kernel,
  nb_id        = 10,
  nb_group     = 2,
  nb_sample    = 30,
  diff_group   = c(0, 2.5),
  var_sample   = 4,
  mu_0         = 40,
  mu_random    = TRUE,
  range_input  = c(0, 100),
  group_labels = c("Control", "Treatment")
)
ids <- unique(data_base$ID[data_base$Group == "Control"])

build_dataset <- function(nb_sample_control, nb_sample_treatment) {
  ctrl <- data_base[data_base$Group == "Control"   & data_base$Sample <= nb_sample_control,   ]
  trt  <- data_base[data_base$Group == "Treatment" & data_base$Sample <= nb_sample_treatment, ]
  rbind(ctrl, trt)
}

scenarios <- list(
  balanced_low   = c(Control = 3,  Treatment = 3),
  treatment_rich = c(Control = 3,  Treatment = 30),
  control_rich   = c(Control = 30, Treatment = 3),
  balanced_high  = c(Control = 30, Treatment = 30)
)
```

## Fitting the kernel once, from a generously-sampled calibration set

A well-specified analysis fits the kernel hyperparameters from data, combining a signal
kernel with a white-noise nugget (exactly as in the introductory vignette). We do this once,
from the full 30-replicate Control group in `data_base`, and reuse the same fitted kernel for
every scenario below  -  this isolates the effect of each group's own sample size on the
*posterior* step from any variation in hyperparameter recovery itself:

```{r}
se_part    <- variance_kernel(variance = 1) * se_kernel(length_scale = 1)
noise_part <- white_noise_kernel(noise = 1)
ker <- se_part + noise_part

control_calib  <- data_base[data_base$Group == "Control", ]
prior_mean_est <- mean(control_calib$Output)
hp_opt  <- BayesOmics::optim_hp(ker, control_calib, prior_mean = prior_mean_est, prior_cov = 1)
ker_opt <- do.call(kupdate, c(list(ker), as.list(hp_opt)))
hp_opt
```

Now build the posterior and draw samples for every (Control, Treatment) sample-size pair,
using that one fitted kernel throughout:

```{r}
scenario_label <- function(s) sprintf("Control=%d, Treatment=%d", s["Control"], s["Treatment"])

scenario_results <- lapply(scenarios, function(s) {
  data      <- build_dataset(s["Control"], s["Treatment"])
  posterior <- multi_posterior_mean(data, ker_opt, mu_0 = prior_mean_est, lambda_0 = 1)
  samples   <- sample_posterior(posterior, n = 2000)
  list(label = scenario_label(s), nb_control = s["Control"], nb_treatment = s["Treatment"],
       posterior = posterior, samples = samples)
})
```

Every scenario above -- including the two strongly unbalanced ones -- runs through
`multi_posterior_mean()` and `sample_posterior()` without any special-casing.

## Each group's posterior reflects only its own sample size

`plot_posterior_mean()` shows the posterior mean at every CpG site for both conditions,
across the four scenarios. Each group's estimates converge more tightly when it has more
replicates  -  regardless of how many replicates the other group has:

```{r, fig.alt = "Posterior mean per CpG site and group, faceted by (Control, Treatment) sample-size scenario; the spread of the Control estimates is controlled only by nb_sample_control, and vice versa"}
plots_mean <- lapply(scenario_results, function(res) {
  plot_posterior_mean(res$samples) +
    ggplot2::ggtitle(res$label)
})
gridExtra::grid.arrange(grobs = plots_mean, nrow = 2, ncol = 2)
```

In the two unbalanced scenarios (`Control=3, Treatment=30` and `Control=30, Treatment=3`),
the group with only 3 replicates shows more dispersion across sites  -  its posterior is
less concentrated  -  while the 30-replicate group is already tightly constrained.
Each group's posterior precision depends only on *its own* replicate count  - 
exactly as the formula says  -  and the comparison between groups remains entirely
valid throughout, because both still share the same `Input` (hence the same
\(\Sigma_{\hat\theta}\)).

## The posterior difference of means still has a well-defined answer

`plot_distrib()`'s difference-of-means density for a single representative site, across the
same four (possibly very unbalanced) scenarios:

```{r, fig.alt = "Posterior density of the difference of means (Control minus Treatment) for one CpG site, faceted by sample-size scenario; well-defined in every scenario including the unbalanced ones"}
focus_id <- ids[3]
plots_diff <- lapply(scenario_results, function(res) {
  plot_distrib(res$samples, group1 = "Control", group2 = "Treatment", id = focus_id,
               show_prob = FALSE) +
    ggtitle(res$label)
})
gridExtra::grid.arrange(grobs = plots_diff, nrow = 2, ncol = 2)
```

```{r}
prob_table <- do.call(rbind, lapply(scenario_results, function(res) {
  diff_samples <- res$samples$Sample[res$samples$ID == focus_id & res$samples$Group == "Control"] -
    res$samples$Sample[res$samples$ID == focus_id & res$samples$Group == "Treatment"]
  data.frame(scenario = res$label,
             `P(Control <= Treatment)` = round(mean(diff_samples <= 0), 3),
             check.names = FALSE)
}))
prob_table
```

The two unbalanced scenarios (`Control=3, Treatment=30` and `Control=30, Treatment=3`) land
at a similar confidence level to each other -- what mostly matters for this comparison is how
much total information is available, not which particular group it came from.

## Seeing the overlap itself, group by group

`plot_posterior_overlap()` overlays the two groups' posterior densities directly on the same
panel and shades their region of intersection -- exactly the area `calculate_group_overlaps()`'s
OVL coefficient measures. This is the most direct way to see an unbalanced comparison still
work cleanly: one curve can be visibly wider than the other, and the shaded overlap is still
well defined.

```{r, fig.alt = "Posterior densities of Control and Treatment overlaid for one CpG site, faceted by sample-size scenario; one curve is visibly wider than the other in the unbalanced scenarios, yet the shaded overlap remains well defined"}
plots_overlap <- lapply(scenario_results, function(res) {
  plot_posterior_overlap(res$samples, group1 = "Control", group2 = "Treatment", id = focus_id) +
    ggtitle(res$label)
})
gridExtra::grid.arrange(grobs = plots_overlap, nrow = 2, ncol = 2)
```

In the unbalanced panels, the narrower curve belongs to whichever group happens to have more
replicates in that scenario -- Treatment in `Control=3, Treatment=30`, Control in
`Control=30, Treatment=3` -- while the wider curve's group is exactly the one with fewer
replicates. The shaded overlap is computed from both curves together regardless.

## The region-wide differential call, across all four scenarios

Finally, `calculate_group_overlaps()`'s single OVL number per scenario, computed from the
full, multivariate posterior profile (all 10 sites at once, accounting for their spatial
correlation) -- this is the closed-form, unequal-scale branch of the overlap formula doing
the work whenever Control and Treatment have different replicate counts:

```{r, fig.alt = "Overlapping coefficient between Control and Treatment posteriors for each of the four (Control, Treatment) sample-size scenarios"}
ovl_df <- do.call(rbind, lapply(scenario_results, function(res) {
  mat <- calculate_group_overlaps(res$posterior)
  data.frame(label = res$label, OVL = mat["Control", "Treatment"])
}))
ovl_df$label <- factor(ovl_df$label, levels = vapply(scenarios, scenario_label, character(1)))
ovl_df
```

All four scenarios -- including the two strongly unbalanced ones -- produce a valid,
well-defined region-wide call. `balanced_low` (3 and 3 replicates) is the least informative,
`balanced_high` (30 and 30) the most; the two unbalanced scenarios sit in between, reflecting
that the comparison is only as precise as its *least* well-sampled group, while still being
perfectly valid to compute.

## Takeaways

The only hard requirement to compare two groups with `calculate_group_overlaps()` is that they share the same `Input` values (same `kernel_key`); their replicate counts (`nb_sample`) can differ freely. Each group's posterior precision depends only on its *own* sample size (`scale = nb_sample + lambda_0` in `multi_posterior_mean()`'s output)  -  adding replicates to one group never changes the other group's uncertainty. When the two groups' scales differ, `calculate_group_overlaps()` automatically uses the exact closed-form overlap for unequal-scale Gaussians (see `?calculate_group_overlaps`)  -  this is what made every unbalanced scenario above work without any special handling. In practice, you do not need to discard "extra" replicates from a richer group, or wait until both groups are equally sampled, to run a valid differential analysis.

## Go further

| Vignette | What you will find |
|---|---|
| `vignette("omics-analysis")` | The complete model walkthrough  -  kernel choice, hyperparameter fitting, posterior computation, and every plot function  -  on both a two-group and a multi-group example. Read this first if you are new to the package. |
| `vignette("troubleshooting")` | Exact error messages you may encounter (non-convergence, mismatched IDs, ill-conditioned matrices, and more), what causes each of them, and the targeted fix. |
