---
title: "Troubleshooting: common errors and how to fix them"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Troubleshooting: common errors and how to fix them}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(keRnel)
library(BayesOmics)
```

This page lists the errors and warnings you are most likely to run into, what they
mean, and how to fix them. Each one is reproduced below with `try()` so the message
matches exactly what you would see in your own console.

## Optimization did not converge

```{r, warning = TRUE}
data <- simu_db(nb_id = 20)
kern <- variance_kernel(variance = 0.01) * se_kernel(length_scale = 0.01)
opt  <- optim_hp(kern, data, prior_mean = 0, prior_cov = 1, max_iter = 1)
```

**Cause:** L-BFGS-B stopped before satisfying its convergence criterion -- usually
because `max_iter` is too low, or the data/kernel combination is poorly conditioned
(`pen_diag` too small, or the kernel's initial construction values far from a good optimum).

**Fix:** increase `max_iter`, try different initial values when constructing `kern`, or increase `pen_diag`.
You don't need `verbose = TRUE` to check: the returned vector always carries the
optimizer's status as attributes:

```{r}
attr(opt, "convergence")  # 0 means success
attr(opt, "value")        # final (negative log-likelihood) objective value
```

## Groups do not share the same set of IDs

```{r, error = TRUE}
bad_results <- list(
  kernels = list(k = diag(2)),
  groups  = list(
    A = list(muk = c(ID_1 = 0, ID_2 = 1), id_to_input = c(ID_1 = 1, ID_2 = 2), kernel_key = "k", scale = 1),
    B = list(muk = c(ID_1 = 0, ID_3 = 1), id_to_input = c(ID_1 = 1, ID_3 = 2), kernel_key = "k", scale = 1)
  )
)
calculate_group_overlaps(bad_results)
```

**Cause:** `calculate_group_overlaps()` compares two groups ID by ID; it requires
every group to be measured on the exact same set of IDs.

**Fix:** restrict the comparison to groups that share all their IDs, or make sure
your input data frame has one row per (ID, Group, Sample) for every ID in every group
before calling `multi_posterior_mean()`.

## Groups do not share the same kernel matrix

**Cause:** two groups have different sets of `Input` values (e.g. different genomic
positions), so their posterior covariances don't derive from the same raw kernel
matrix. The closed-form overlap formula used by `calculate_group_overlaps()` only
applies when both groups share one kernel matrix, scaled differently by `scale`.

**Fix:** make sure every group in your data is measured at the same `Input` values
for the same `ID`s -- `simu_db()` and `simu_db_kernel()` do this by construction. If
your real data genuinely has different `Input` values per group, you cannot compare
those two groups with `calculate_group_overlaps()` directly.

## Each ID must map to a single Input value within its group

```{r, error = TRUE}
data_bad <- data.frame(
  ID     = c("ID_1", "ID_1"),
  Group  = "A",
  Output = c(0, 0),
  Input  = c(5, 6)  # same ID, two different Input values
)
multi_posterior_mean(data_bad, se_kernel(length_scale = 1))
```

**Cause:** within a single group, `ID` must uniquely determine `Input` (e.g. one CpG
site cannot sit at two different genomic positions in the same condition). This
usually comes from a data-cleaning issue -- duplicated rows, or an `ID` reused across
two physically different positions.

**Fix:** check for duplicate/inconsistent `(Group, ID)` rows in your input data
frame with `dplyr::distinct(data, Group, ID, Input)` and resolve the inconsistency
before calling `multi_posterior_mean()`.

## Group has IDs with different numbers of observations

```{r, error = TRUE}
data_uneven <- data.frame(
  ID     = c("ID_1", "ID_1", "ID_2"),
  Group  = "A",
  Sample = c(1, 2, 1),
  Output = c(0, 1, 0),
  Input  = c(5, 5, 6)
)
multi_posterior_mean(data_uneven, se_kernel(length_scale = 1))
```

**Cause:** `multi_posterior_mean()` uses a single observation count per group (the
posterior `scale` is `n_obs + lambda_0`); it requires every ID within a group to have
the same number of replicate samples.

**Fix:** make sure every ID in a group has the same number of `Sample` rows -- e.g.
by filling in missing replicates or subsetting to a balanced design. See
`vignette("sample_size_scenarios")` for designs with different replicate counts
*across* groups (which is supported), as opposed to across IDs *within* the same
group (which is not).

## Matrix could not be made positive-definite

**Cause:** internally, `chol_inv_jitter()` adds a small jitter to a covariance
matrix's diagonal to make it invertible, increasing the jitter geometrically up to
20 times. If it still fails, the matrix is likely severely ill-conditioned -- e.g.
near-duplicate `Input` values combined with a very large kernel length scale.

**Fix:** increase `pen_diag`, reduce the kernel's length scale, or check for
near-duplicate `Input` values in your data.

## Empty groups

**Cause:** a `Group` value in your data has no `ID` or no `Input` value associated
with it -- usually an artefact of filtering the data frame before calling
`multi_posterior_mean()`.

**Fix:** drop unused factor levels / filter out the offending rows before fitting,
e.g. `data <- droplevels(data[data$Group %in% kept_groups, ])`.

## Go further

| Vignette | What you will find |
|---|---|
| `vignette("omics-analysis")` | The complete model walkthrough  -  kernel choice, hyperparameter fitting, posterior computation, and every plot function  -  on both a two-group and a multi-group example. |
| `vignette("sample_size_scenarios")` | How unbalanced designs (different replicate counts per group) affect posterior width and the final OVL, and why the comparison remains valid even with strongly unequal counts. |
